---
title: "Factorial Analysis of IPIP Big Five Personality Data"
format: html
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Imports

```{r}
library(psych)       # factor analysis
library(ggplot2)     # data visualisation
library(pheatmap)    # heatmaps
library(tidyr)       # data reshaping
library(dplyr)       # data manipulation
library(gridExtra)   # combining multiple plots

options(digits = 5)
```

# Data Cleaning

```{r}
data_raw <- read.csv("./IPIP-FFM-data-8Nov2018/data-final.csv", sep = "\t", na.strings = c("", " ", "NaN", "NA","NULL"))
```

```{r}
# Select columns to only keep the 50 items that measure the 5 personality traits.
# See codebook.txt for more details on all the columns.
data <- data_raw[, 1:50]

# Remove rows where IPC != 1, as these could be multiple submissions of the same person.
# The total number of rows in the original dataset is 1 015 341, and 696 845 after cleaning.
cat("Number of lines removed to avoid multiple submission:", sum(data_raw$IPC != 1), "\n")
data <- data[data_raw$IPC == 1, ]

# Remove rows with missing data
n_before <- nrow(data)
data <- data[complete.cases(data), ]
n_after  <- nrow(data)
cat("Number of lines removed to avoid missing data:", n_before - n_after, "\n")
cat("Number of lines :", n_after)
data_mat <- as.matrix(data)
head(data)
```

```{r}
# Computing polychoric correlations on large datasets is computationally heavy.
# We check if the matrix has already been saved to avoid recalculating it every time.

pmat_path <- "polychoric_matrix_IPIP-50.rds"

if (file.exists(pmat_path)) {
  cat("Loading saved polychoric matrix...\n")
  poly_mat <- readRDS(pmat_path)
} else {
  cat("Computing polychoric matrix (this may take a while)...\n")
  poly_mat_obj <- polychoric(data_mat, ML=FALSE) #ML=FALSE (default value) to use Olsson's two step procedure 
  poly_mat <- poly_mat_obj$rho
  
  # Save the matrix to a file for future runs
  saveRDS(poly_mat, pmat_path)
  cat("Polychoric matrix saved to", pmat_path, "\n")
}
```

# Factor Analysis

We perform factor analysis on the data to reduce the dimensionality of
the data. In psychological terms, it means finding the underlying
fundamental personality traits that (partially) explain the responses to
the questions.

## Note on Rotation and Interfactorial Correlations

We apply factor analysis to the data, to extract the Big Five
personality traits, with varimax rotation to make the traits more
interpretable.

We could have used non-orthogonal rotation methods, like oblimin, which
allow for correlated factors. It is commonly recommended to use oblique
rotation in psychometric analysis, as said in Fabrigar et al.
(<https://doi.org/10.1037/1082-989X.4.3.272>):

> *For many constructs examined in psychology (e.g., mental abilities,
> personality traits, attitudes), there is substantial theoretical and
> empirical basis for expecting these constructs (or dimensions of these
> constructs) to be correlated with one another. Therefore, oblique
> rotations provide a more accurate and realistic representation of how
> constructs are likely to be related to one another.*

However, our goal here is not to explain the correlations between
factors, but to perform clustering on the factors, and we fear that
oblique rotation will deform the space in a way that will influence the
clustering. Additionally, Costello and Osborne
(<https://doi.org/10.1037/1082-989X.4.3.272>) suggest that orthogonal
and oblique rotation often produce similar results, so we will not lose
much interpretability by using varimax rotation. The results given in
appendix B of this notebook agree with this statement.

## Note on the Factor Analysis Extraction Method

`psych::fa()` with `fm="minres"` uses the ULS/minres method (also called
the iterated principal factor method) to perform factor analysis, which
does not assume that factors are normally distributed, contrary to the
maximum likelihood method.

Since the assumption that factors follow a Gaussian distribution
directly contradicts the assumption of the Gaussian mixture model we
will use later, we think it is preferable to use the ULS method.
However, this does not seem to bother Gerlach, M., Farb, B., Revelle, W.
et al. (<https://doi.org/10.1038/s41562-018-0419-z>), who use the ML
method before using GMM. For the sake of comparison, we also perform
factor analysis with the ML method in appendix A of this notebook, and
the results are very similar to those given by the ULS method.

## How Many Factors to Extract?

A common method to determine the number of factors to extract is to look
at the scree plot, which is a plot of the eigenvalues of the correlation
matrix of the data. The eigenvalues represent the amount of variance
explained by each factor. The scree plot shows a clear elbow at 5
factors, which suggests that we should extract 5 factors. This is
consistent with the fact that we are trying to extract the Big Five
personality traits.

```{r}
#| warning: false
# The ULS method has several names: minres, iterated principal factor method,
# or unweighted least squares method.
fa_result <- fa(poly_mat, nfactors = 5, rotate = "varimax", fm = "minres", n.obs = nrow(data_mat))

data_fa <- as.data.frame(factor.scores(data_mat, fa_result)$scores)

# Eigenvalues of the original (unreduced) correlation matrix
eigenvalues <- fa_result$values
cat("Eigenvalues:", eigenvalues, "\n")

# Plot the ten first eigenvalues in a scree plot
top_10_eigenvalues <- eigenvalues[1:10]
scree_df <- data.frame(factor = 1:10, eigenvalue = top_10_eigenvalues)

ggplot(scree_df, aes(x = factor, y = eigenvalue)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(color = "blue", size = 3) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Scree Plot", x = "Factor Number", y = "Eigenvalue") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

Looking at the scree plot, it is not clear that we should extract five
factors; the solution with six factors also seems plausible.

```{r}
#| warning: false
# Run MAP and VSS test on the polychoric correlation matrix
n_of_factors_diagnostic <- VSS(poly_mat, n.obs = nrow(data_mat), n = 10, plot = FALSE)
summary(n_of_factors_diagnostic)
```

```{r}
# Plot the weights of the 50 items on the 5 personality traits as a heatmap.
# NOTE: The factor order below is inferred by inspecting which group of items
# loads most strongly on each factor (see the loading heatmap).
trait_names <- c("Extraversion", "Neuroticism", "Agreeableness",
                 "Conscientiousness", "Openness")

data_fa  <- data_fa[, c(2,1,3,5,4)]
fa_result$loadings <- fa_result$loadings[, c(2,1,3,5,4)]

# fa$loadings is items × factors; transpose to get traits × items
weights <- t(as.matrix(fa_result$loadings))


rownames(weights) <- trait_names

coolwarm <- colorRampPalette(c("#3B4CC0", "white", "#B40426"))(200)
lim      <- max(abs(weights))
breaks   <- seq(-lim, lim, length.out = 201)

pheatmap(
  weights,
  color        = coolwarm,
  breaks       = breaks,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main         = "Weights of the 50 items on the 5 personality traits",
  fontsize      = 8,
  fontsize_row  = 9
)
```

```{r}
#| fig-width: 15
#| fig-height: 4
# From the previous plot, we can infer that factors are ordered as listed in trait_names.
# Assign the Big Five trait names as column names.
colnames(data_fa) <- trait_names


data_fa_long <- pivot_longer(data_fa, cols = everything(),
                             names_to = "trait", values_to = "score")
data_fa_long$trait <- factor(data_fa_long$trait)

ggplot(data_fa_long, aes(x = score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60, alpha = 0.6, color = NA) +
  geom_density(color = "darkred", linewidth = 0.8) +
  facet_wrap(~trait, nrow = 1, scales = "free") +
  theme_minimal(base_size = 11) +
  labs(x = "Score", y = "Density")
```

# Appendix A: Comparison of ULS vs ML Extraction Methods for Factor Analysis

```{r}
#| warning: false
#| fig-height: 4
#| fig-width: 15
# Perform factor analysis with the maximum likelihood (ML) method,
# for comparison with the ULS method.
fa_ml     <- fa(poly_mat, nfactors = 5, rotate = "varimax", fm = "ml", n.obs = nrow(data_mat))
data_fa_ml <- as.data.frame(factor.scores(data_mat, fa_ml)$scores)

# Exchange column to properly match with traits
data_fa_ml  <- data_fa_ml[, c(1, 2, 3, 5, 4)]
fa_ml$loadings <- fa_ml$loadings[, c(1, 2, 3, 5, 4)]

colnames(data_fa_ml) <- trait_names
loadings_ml <- as.matrix(fa_ml$loadings)

weights_ml <- t(loadings_ml)

# Loading matrix comparison plots
breaks_full <- seq(-1, 1, length.out = 201)

p1 <- pheatmap(weights_ml, color = coolwarm, breaks = breaks_full,
               cluster_rows = FALSE, cluster_cols = FALSE,
               main = "Factor loading (ML Method)",
               fontsize = 7, fontsize_row = 8, silent = TRUE)
p2 <- pheatmap(weights,    color = coolwarm, breaks = breaks_full,
               cluster_rows = FALSE, cluster_cols = FALSE,
               main = "Factor loading (ULS Method)",
               fontsize = 7, fontsize_row = 8, silent = TRUE)
diff_ml <- unclass(weights)- unclass(weights_ml) #unclass to avoid the rounding of "loadings" class
p3 <- pheatmap(diff_ml,    color = coolwarm, breaks = breaks_full,
               cluster_rows = FALSE, cluster_cols = FALSE,
               main = "Difference (ML - ULS)",
               fontsize = 7, fontsize_row = 8, silent = TRUE)

grid.arrange(p1$gtable, p2$gtable, p3$gtable, ncol = 3,
             widths = c(1, 1, 1))

cat("Difference between the weights of the ML method and the ULS method:\n")
diff_ml
```

```{r}
#| fig-height: 4
#| fig-width: 15
# Histogram comparison: ML vs ULS distributions for each trait

data_compare <- bind_rows(
  pivot_longer(data_fa_ml, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "ML"),
  pivot_longer(data_fa, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "ULS")
)
data_compare$trait <- factor(data_compare$trait)

ggplot(data_compare, aes(x = score, color = method, fill = method)) +
  
  geom_density(alpha = 0.4, linewidth = 0.8) +
  
  facet_wrap(~trait, nrow = 1, scales = "free") +
  
  scale_fill_manual(values  = c("ML" = "royalblue", "ULS" = "darkred")) +
  scale_color_manual(values = c("ML" = "royalblue", "ULS" = "darkred")) +
  
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom") +
  labs(x = "Score", y = "Density")
```

We can see that the loadings and distributions of the factors with the
ULS method are very similar to those given by the factor analysis with
the ML method, which suggests that the choice of extraction method does
not have a strong influence on the results. However, we think it is
preferable to use the ULS method, since it does not assume that factors
are normally distributed, which is more consistent with the assumptions
of the Gaussian mixture model we will use later.

# Appendix B: Comparison of Varimax vs Oblimin Rotation Methods for Factor Analysis

In the main part of the notebook, we use orthogonal rotation (varimax)
to make the factors more interpretable. However, we could have used
oblique (oblimin) rotation, which allows for correlated factors. We will
see that the results given by the oblimin rotation are very similar to
those given by the varimax rotation, which suggests that the choice of
rotation method does not have a strong influence on the results.

```{r}
#| warning: false
#| fig-height: 4
#| fig-width: 15
# Perform factor analysis with the oblimin rotation method,
# for comparison with the varimax rotation used above.
fa_obli      <- fa(poly_mat, nfactors = 5, rotate = "oblimin", fm = "minres", n.obs = nrow(data_mat))
data_fa_obli <- as.data.frame(factor.scores(data_mat, fa_obli)$scores)

# Exchange column to properly match with traits
data_fa_obli  <- data_fa_obli[, c(1, 2, 3, 5, 4)]
fa_obli$loadings <- fa_obli$loadings[, c(1, 2, 3, 5, 4)]

loadings_obli  <- as.matrix(fa_obli$loadings)

# From our (not shown) observations, factors are ordered as follows:
colnames(data_fa_obli) <- trait_names

weights_obli <- t(loadings_obli)
rownames(weights_obli) <- trait_names

# Loading matrix comparison plots
p_obli <- pheatmap(weights_obli, color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Factor loading (Oblique Rotation)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)
p_vari <- pheatmap(weights,      color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Factor loading (Orthogonal Rotation)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)
diff_obli <- unclass(weights_obli) - unclass(weights)
p_diff <- pheatmap(diff_obli,   color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Difference (Oblique - Orthogonal)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)

grid.arrange(p_obli$gtable, p_vari$gtable, p_diff$gtable, ncol = 3,
             widths = c(1, 1, 1))

cat("Difference between the weights of varimax and oblimin rotation:\n")
diff_obli
```

## Correlation Between the Factors Obtained with the Oblimin Rotation

Unlike the factors obtained with the varimax rotation, the factors
obtained with the oblimin rotation are correlated, as expected. However,
the correlations between factors are not very strong, which suggests
that the factors are still relatively independent from each other. This
is consistent with the fact that the results given by the oblimin
rotation are very similar to those given by the varimax rotation. We
give here a comparison of the correlation matrix we found with the one
given by Van der Linden et al. in their 2010 meta-analysis
(<https://doi.org/10.1016/j.jrp.2010.03.003>).

```{r}
#| fig-height: 4
#| fig-width: 11
# Correlation matrix of the 5 personality traits obtained with the oblique rotation
corr_obli <- cor(data_fa_obli)

# (Uncorrected) correlation matrix from the meta-analysis of Van der Linden et al. (2010)
corr_meta <- matrix(c(
   1.00, -0.26,  0.18,  0.21,  0.31,
  -0.26,  1.00, -0.26, -0.32, -0.12,
   0.18, -0.26,  1.00,  0.31,  0.14,
   0.21, -0.32,  0.31,  1.00,  0.14,
   0.31, -0.12,  0.14,  0.14,  1.00
), nrow = 5, byrow = TRUE,
   dimnames = list(trait_names, trait_names))

sd_meta <- matrix(c(
  0.00, 0.11, 0.15, 0.15, 0.12,
  0.11, 0.00, 0.14, 0.18, 0.12,
  0.15, 0.14, 0.00, 0.14, 0.12,
  0.15, 0.18, 0.14, 0.00, 0.15,
  0.12, 0.12, 0.12, 0.15, 0.00
), nrow = 5, byrow = TRUE)

# Build annotation matrices
annot_obli <- matrix(sprintf("%.2f", corr_obli), nrow = 5,
                     dimnames = dimnames(corr_obli))
annot_meta <- matrix("", nrow = 5, ncol = 5, dimnames = dimnames(corr_meta))
for (i in 1:5) {
  for (j in 1:5) {
    if (i == j) {
      annot_meta[i, j] <- sprintf("%.2f", corr_meta[i, j])
    } else {
      annot_meta[i, j] <- sprintf("%.2f \n(+/-%.2f)", corr_meta[i, j], sd_meta[i, j])
    }
  }
}

# Show the two correlation matrices side by side for comparison
p_corr_obli <- pheatmap(
  corr_obli, color = coolwarm, breaks = breaks_full,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = annot_obli, number_format = "%.2f",
  main = "Correlation Matrix (Oblique Rotation)",
  fontsize = 9, silent = TRUE
)
p_corr_meta <- pheatmap(
  corr_meta, color = coolwarm, breaks = breaks_full,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = annot_meta,
  main = "Correlation Matrix (Meta-Analysis)",
  fontsize = 9, silent = TRUE
)

grid.arrange(p_corr_obli$gtable, p_corr_meta$gtable, ncol = 2,
             widths = c(1, 1))
```

```{r}
#| fig-height: 4
#| fig-width: 11
# Histogram comparison: oblimin vs varimax distributions for each trait
data_compare_obli <- bind_rows(
  pivot_longer(data_fa_obli, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "oblimin (oblique)"),
  pivot_longer(data_fa, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "varimax (orthogonal)")
)
data_compare_obli$trait <- factor(data_compare_obli$trait)

ggplot(data_compare_obli, aes(x = score, color = method, fill = method)) +
  
  geom_density(alpha = 0.4, linewidth = 0.8) +
  
  facet_wrap(~trait, nrow = 1, scales = "free") +
  
  scale_fill_manual(values  = c("oblimin" = "royalblue", "varimax" = "darkred")) +
  scale_color_manual(values = c("oblimin" = "royalblue", "varimax" = "darkred")) +
  
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom") +
  labs(x = "Score", y = "Density")
```

# Appendix C: Pearson vs polychoric matrix for factor extraction

## Polychoric correlation matrix

### The trouble with Likert scales

Exploratory factor analysis is based on a correlation matrix: from a
correlation matrix, it infers the underlying factors that explain the
correlations between the items. The core assumption of our study is that
the responses to the questionnaire are strongly determined by some
underlying *continuous* factors. So it seems intuitive that each person
has an "internal" *continuous* response to each question that is a
linear combination of the underlying (continuous) factors plus a random
noise. This "internal" response is then discretized into an ordinal
Likert scale when answering the questionnaire. In a perfect world, we
would have direct access to these "internal" continuous responses, which
are more reflective of the personality of the participant, and then we
would perform the factor analysis on the correlations of these
"internal" continuous responses. But we only have access to the Likert
scales, which correlation underestimate the "true" underlying
correlation, as we can see in the following theorem:

> **Theorem:** If $(Z_1, Z_2)$ have a joint bivariate normal
> distribution with correlation coefficient $\rho$, then for any
> Borel-measurable functions $f$ and $g$ such that $f(Z_1)$ and $g(Z_2)$
> possess finite variances, the absolute correlation between the
> transformed variables is bounded by the absolute value of $\rho$:
> $$|\text{corr}(f(Z_1), g(Z_2))| \leq |\rho|$$ The equality only
> happens when $f=g=id$.

> **Proof:** The proof rely on the fact that we can express $L^2$ random
> variables (w.r.t. Gaussian measure) as series of standardized
> Hermite-Tchebycheff polynomial, which are orthogonal w.r.t. Gaussian
> measure. See [H.O. Lancaster(1957)](https://doi.org/10.2307/2333274)
> for the full proof.

Now if we assume that $Z_1$ and $Z_2$ are Gaussian, we can define $L_1$
and $L_2$ as the observed ordinal Likert variables. These discrete
variables are created by partitioning the continuous variables $Z_1$ and
$Z_2$ using a set of thresholds (e.g., $L_1 = i$ if
$\alpha_{i-1} < Z_1 \leq \alpha_i$, and $L_2 = j$ if
$\beta_{j-1} < Z_2 \leq \beta_j$). Since $L_1$ and $L_2$ are created
through step functions of the underlying continuous variables (i.e.,
$L_1 = f(Z_1)$ and $L_2 = g(Z_2)$), it rigorously follows that the
correlation between the discrete responses is strictly bounded by the
correlation of the "internal" continuous variables:

$$|\text{corr}(L_1, L_2)| < |\rho|$$

This is why Likert-scale questionnaires tend to underestimate the "true"
correlation between the questions. From this example, it seems very
plausible that the naive use of Pearson correlation on Likert scale can
give poor results. One way of addressing this issue is to use a
polychoric correlation matrix.

### Polychoric correlation estimation

Given what we have said above, if we use the Pearson correlation matrix,
we will obtain an estimation of the correlation of the Likert discrete
variables, not the correlation of the underlying continuous ones. So we
need a way to infer the correlation of the underlying continuous
variables given their discretized versions. This is exactly what
polychoric correlation is supposed to do.

To compute the polychoric correlation, we assume that the underlying
continuous latent variables $(Z_1, Z_2)$ follow a standard bivariate
normal distribution with correlation $\rho$. The theoretical probability
$\pi_{ij}$ of an observation falling into the $i$-th category for $L_1$
and the $j$-th category for $L_2$ is the following:

$$\pi_{ij} = \Phi_2^{}(\alpha_i, \beta_j; \rho) - \Phi_2(\alpha_{i-1}, \beta_j; \rho) - \Phi_2(\alpha_i, \beta_{j-1}; \rho) + \Phi_2(\alpha_{i-1}, \beta_{j-1}; \rho)$$

where $\Phi_2$ is the cumulative distribution function (CDF) of the
standard bivariate normal distribution.

To estimate these parameters from an observed dataset where $n_{ij}$
represents the frequency of responses in cell $(i, j)$, we use Maximum
Likelihood Estimation (MLE). The log-likelihood function to maximize is
defined as:

$$\ell(\rho, \mathbf{\alpha}, \mathbf{\beta}) = \sum_{i} \sum_{j} n_{ij} \ln[\pi_{ij}(\rho, \alpha, \beta)]$$

**Olsson's Two-Step Method** Maximizing this log-likelihood for all
parameters ($\rho$, $\alpha$, $\beta$) simultaneously is computationally
expensive. To solve this, the standard approach relies on [Olsson's
(1979)](https://doi.org/10.1007/BF02296207) two-step procedure, which
consists of first estimating the thresholds $\alpha$ and $\beta$ using
only the marginal distributions, and then estimating the correlation
coefficient $\rho$. Despite the theoretical non-optimality of this
procedure, the gain in computational efficiency outweighs the loss in
precision, as reported in Olsson's article. However, it is still
possible to use a full maximum likelihood estimation for maximum
statistical efficiency. In this code, we will use Olsson's two step
method (parameter `ML=FALSE` in the `polychoric` function from `psych`
package).

### Empirical evidence of the superiority of polychoric correlation matrix in EFA

The methodological importance of this choice is highlighted by
Holgado–Tello et al (<https://doi.org/10.1007/s11135-008-9190-y> ). The
authors demonstrate that when data are ordinal, polychoric correlations
provide a more accurate and stable reproduction of the original factor
structure compared to Pearson correlations, which can lead to biased
loadings and incorrect factor counts.

```{r}
#| warning: false
#| message: false
#| fig-height: 4
#| fig-width: 15

# Perform factor analysis with the oblimin rotation method using Pearson correlations.
# NOTE: The default was set to polychoric at the beginning of the notebook for all main analyses.
fa_pearson      <- fa(data_mat, nfactors = 5, rotate = "oblimin", fm = "minres",
                   cor = "cor", scores = "regression")
data_fa_pearson <- as.data.frame(fa_pearson$scores)
loadings_pearson  <- as.matrix(fa_pearson$loadings)

# Apply the trait names to the columns and rows
colnames(data_fa_pearson) <- trait_names

weights_pearson <- t(loadings_pearson)
rownames(weights_pearson) <- trait_names

# Loading matrix comparison plots
p_poly <- pheatmap(weights_obli, color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Factor loading (Polychoric Oblimin)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)

p_pearson <- pheatmap(weights_pearson, color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Factor loading (Pearson Oblimin)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)

# Calculate difference using unclass() to prevent the "loadings" rounding issue
diff_pearson <- unclass(weights_pearson) - unclass(weights_obli)

p_diff <- pheatmap(diff_pearson,   color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Difference (Pearson - Polychoric)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)

grid.arrange(p_poly$gtable, p_pearson$gtable, p_diff$gtable, ncol = 3,
             widths = c(1, 1, 1))

cat("Difference between the weights of Pearson and Polychoric:\n")
print(round(diff_pearson, 5))
```

```{r}
#| fig-height: 4
#| fig-width: 15
# Histogram comparison: Pearson vs Polychoric correlation matrix for factor analysis for each trait

data_compare_pearson <- bind_rows(
  pivot_longer(data_fa_pearson, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "Pearson"),
  pivot_longer(data_fa_obli, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "Polychoric")
)

# Convert trait to factor to control facet order if needed
data_compare_pearson$trait <- factor(data_compare_pearson$trait)

ggplot(data_compare_pearson, aes(x = score, color = method, fill = method)) +
  
  geom_density(alpha = 0.4, linewidth = 0.8) +
  
  facet_wrap(~trait, nrow = 1, scales = "free") +
  
  scale_fill_manual(values  = c("Pearson" = "royalblue", "Polychoric" = "darkred")) +
  scale_color_manual(values = c("Pearson" = "royalblue", "Polychoric" = "darkred")) +
  
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom") +
  labs(x = "Score", y = "Density")
```

```{r}
#| fig-height: 5
#| fig-width: 16
#| warning: false

corr_poly <- cor(data_fa_obli)

corr_pearson <- cor(data_fa_pearson)

annot_poly <- matrix(sprintf("%.2f", corr_poly), nrow = 5, 
                     dimnames = dimnames(corr_poly))
annot_pearson <- matrix(sprintf("%.2f", corr_pearson), nrow = 5, 
                        dimnames = dimnames(corr_pearson))

p_corr_poly <- pheatmap(
  corr_poly, color = coolwarm, breaks = breaks_full,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = annot_poly, number_format = "%.2f",
  main = "Factor Correlation (Polychoric Oblimin)",
  fontsize = 9, silent = TRUE
)

p_corr_pearson <- pheatmap(
  corr_pearson, color = coolwarm, breaks = breaks_full,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = annot_pearson, number_format = "%.2f",
  main = "Factor Correlation (Pearson Oblimin)",
  fontsize = 9, silent = TRUE
)

p_corr_meta_comp <- pheatmap(
  corr_meta, color = coolwarm, breaks = breaks_full,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = annot_meta,
  main = "Factor Correlation (Meta-Analysis)",
  fontsize = 9, silent = TRUE
)

grid.arrange(p_corr_poly$gtable, p_corr_pearson$gtable, p_corr_meta_comp$gtable, 
             ncol = 3, widths = c(1, 1, 1))
```

```{r}
#| fig-height: 5
#| fig-width: 16
#| warning: false

pearson_mat <- cor(data_mat)


diff_item_mat <- poly_mat - pearson_mat

lim_diff <- max(abs(diff_item_mat))
breaks_diff <- seq(-lim_diff, lim_diff, length.out = 201)

p_item_pearson <- pheatmap(
  pearson_mat, 
  color = coolwarm, 
  breaks = breaks_full,
  cluster_rows = FALSE, 
  cluster_cols = FALSE,
  main = "Item Correlation (Pearson)",
  show_rownames = FALSE, 
  show_colnames = FALSE, 
  silent = TRUE
)

p_item_poly <- pheatmap(
  poly_mat, 
  color = coolwarm, 
  breaks = breaks_full,
  cluster_rows = FALSE, 
  cluster_cols = FALSE,
  main = "Item Correlation (Polychoric)",
  show_rownames = FALSE, 
  show_colnames = FALSE, 
  silent = TRUE
)

p_item_diff <- pheatmap(
  diff_item_mat, 
  color = coolwarm, 
  breaks = breaks_diff,
  cluster_rows = FALSE, 
  cluster_cols = FALSE,
  main = "Difference (Polychoric - Pearson)",
  show_rownames = FALSE, 
  show_colnames = FALSE, 
  silent = TRUE
)

# Display the heatmaps side by side
grid.arrange(
  p_item_pearson$gtable, 
  p_item_poly$gtable, 
  p_item_diff$gtable, 
  ncol = 3, 
  widths = c(1, 1, 1)
)

# Print a numerical summary of the differences in the console
cat("Numerical summary of the differences (Polychoric - Pearson) between items:\n")
summary(as.vector(diff_item_mat))
```

As we have said previously, the previous plot clearly indicates that
Pearson correlation tends to underestimate the true correlation between
*questions* (not factor).
